---
title: "First Steps with R and RStudio"
author: Matt Steele
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: yes
  html_document:
    toc: yes
    toc-depth: 1
keep-latex: true
---

# Continue Learning with these Resources

### R download and documentation
  
```{r Official R and CRAN, eval= FALSE}

browseURL("https://cran.r-project.org/")

```
  
### RStudio download and documentation
  
```{r Rstudio, eval=FALSE}

browseURL("https://rstudio.com/")

```
  
### Learning Platform
  
```{r eval=FALSE}

browseURL("https://databases.lib.wvu.edu/connect/1540334373")

```
  
### R Programming for Statistics and Data Science
  
```{r eval=FALSE}

browseURL("https://libwvu.on.worldcat.org/oclc/1062397089")

```
  
---

# Getting Started

### Start a New File

You can create new files (R Scripts or RMarkdown files) that allow you to document your code and the outputs of your code.

  - File > New File > **Type of Document you want to create**

### Set your Preferences 
  
You can set your preferences in RStudio.

  - Tools > Global Options

---

# Setting your working directory

The working directory in R is the folder where you are working. Hence, it’s the place (the environment) where you have to store your files of your project in order to load them or where your R objects will be saved.

### Function: getwd()

See the current directory you are in
      
```{r Getwd help, eval=FALSE}

#HELP FOR getwd()

?getwd

```
    
```{r run getwd, eval = FALSE}

# USING THE getwd() function

getwd()

```

### Function: setwd(**path**)

You can set your working directory in RStudio by going to

  - Session > Set Working Directory > Choose Directory
  - Choose the folder you want to work out of
  
After you set your working directory, save the path by copy and pasting the file path from the console area into the source area using the setwd() function

```{r setwd,  eval = FALSE}

# Remember to save your working directory path to your script or markdown file

setwd("C:/Users/Matt/Documents/RWorkshop Development/workshop_firstSteps_R-main")

```

---

# Overview of IDE in RStudio

RStudio is an Integrated Development Environment (IDE) that allows you to save you code, store your variables and environments and view outputs.

### Console Pane
  
pane on the left-hand side of the screen
    
  - This area is where your code is run and where outputs are displayed
    
### Source Pane
  
this pane is opened when you create or open a markdown or script file.
    
  - This area is where you can create code in script or markdown files
  - These files can be a saved and accessed at any point
    
### Environment Pane
  
view functions, objects, and data sets that are stored here
    
  - Your environment can be saved and accessed at any point
  - Save your environment to your working directory
    
---

# Tips and Tricks

  1. Keyboard Shortcuts
    - Tools > Keyboard Shortcut Tricks
    - Run Code: CTRL + ENTER (PC)
    - Run Code: CMD + RETURN (MAC)
    - Assignment Operator: ALT + - (<-)
  2. Help
    + You can access the help tab in the button right.
    + Contains overviews of topics and functions.
    + You can also call it with code
      + Example: ?getwd
      + Example: help("read.table")
  3. TAB Button
    + the TAB button auto-suggests
      + Functions
      + Arguments
      + Files in WD
      + Values
  4. Commenting
    + **#** - single line
    + **''' text '''** - multiline
    + In script files or in coding areas, you can add text
      + This text will be ignored when the code is run
      + It is used for documentation purposes
    
---

# Getting Started with Coding

### Using Code with Calculations

R can be used to perform calculations
  
#### Example Calculations
      
```{r arthimetic, eval=FALSE}

# Addition:
  129 + 3483
  
  # Subtraction:
  23693 - 4536
  
  # Multiplication:
  23 * 45
  
  # Division:
  51/3
  
  # Exponents:
  2^4
  
  # Logarithms (base e):
  log(100)

```
      
### Inputing Ranged Values
  
You can also call a range of values using (:).

### Example Range
  
```{r range,  eval = FALSE}

1:50

```
---

# Built-In Functions and Arguments

Like most computer software, R allows you to run commands. Commands in R are referred to as functions. There are several built-in functions in base R, however, when you install and call a new package, you will have access to more functions that you can use.

### Base R Built-in Functions

``` {r Base R documentation, eval = FALSE}

browseURL("https://stat.ethz.ch/R-manual/R-devel/library/base/html/00Index.html")

```

#### Example Function
  
```{r function, eval=FALSE}

sample(1:500)

```
  
### Function's Arguments 

values and parameters that are acted on by the function.

  - When a function is invoked, you pass a value to the argument. 
  - Arguments are optional; that is, a function may contain no arguments.
  - Functions can have arguments with default values.
  - The arguments of a function can be called in the order that they appear in their documentation
    
#### Example Argument

```{r argument, eval = FALSE}

sample(1:500, size = 30, replace = TRUE)

```

---

# Objects

Objects allow you to store information for future recall

  - Things that can be objects:
    + single digit
    + a character
    + a vector
    + an imported data frame (e.g.csv)
    + a created data frame
      
### Assignment Operator (<-)

The assignment operator allows you to create an object

  - *syntax: objectName <- stored_value*
    
```{r object, eval=FALSE}

a <- 3

b <- 4

a + b

```
  
### Naming Conventions
  
Objects must start with a lowercase letter and cannot start with a captial letter. Ideally, you should not name your object with the name of a function, to avoid confusion.

  - If your object is using two words or more
    + Example 1 - longer.name
    + Example 2 - longer_name
    + Example 3 - longerName

```{r naming conventions, eval=FALSE}

  first.example <- sample(3:15, size = 1)
    
  second_example <- round(2.471, digits = 2)
    
  thirdExample <- mean(500:4000)
  
  first.example + second_example / thirdExample
  
```

---

# Data Values

### Double 

regular numbers (large small, positive, negative, with digits after the decimal, or without)

```{r double,  eval=FALSE}

  str(400)
 

```

### Integer

positive whole number with nothing after the decimal. Denoted using L

```{r integer, eval=FALSE}

str(400L)

```
  
### Character
  
non-numeric data to be interpreted as text. Denoted using ("")

```{r character, eval=FALSE}

str("Hello World")

```
  
### Factor
  
data objects which are used to categorize the data and store it as levels. 

```{r factor, eval=FALSE}

# Line 333

tarantino <- read.csv ("https://raw.githubusercontent.com/fivethirtyeight/data/master/tarantino/tarantino.csv", stringsAsFactors = T)

str(tarantino$word)

```

### Boolean

boolean data objects that denote TRUE or FALSE. Denoted with capital letters.

```{r boolean, eval=FALSE}

str(TRUE)

```

---

# Packages

R is open source code after it's initial development people began adding to it with packages. An R Package is something that you can plug into RStudio to extend the basic functionality that is built in with R. One of the reasons that R has become so popular is because it has this rich ecosystem of packages that really make R a comprehensive platform for data science.

### Install Packages
  
You must install a package before you can call it. But you only need to install it one time
       
```{r help install packages, eval=FALSE}

help("install.packages")

```
   
```{r install tidyverse, eval=FALSE}

#INSTALL THE TIDYVERSE PACKAGE

install.packages("tidyverse")

```

### Load Packages
  
For every new session, you must load it to use the package's functions

```{r library, eval=FALSE}

# Help for library function

help("library")

```

```{r load tidyverse, eval=FALSE}

library(tidyverse) # you need to call the library during each session

```

---

# Creating Vectors
  
one-dimensional sequence of data elements

  - use *c(... , ...)* To make a vector
  - separate elements using *, (comma)*
  - syntax: *vectorSyntax <- c(object1, object2, object3)*
    
```{r vector, eval=FALSE}

vec_one <- c(1,2,10)
vec_two <- c(5,6,7)
vec_three <- c("Morgantown", "Charleston", "Huntington")

```

```{r vector mean, eval=FALSE}

mean(vec_one)

```

### RBIND and CBIND 

combine vectors to create two-dimensional arrays.Combinations with the same data value will create a *matrix*. Combinations with different values will create a *data frame*.
      
```{r cbind, eval=FALSE}

cbind(vec_one, vec_two, vec_three)

```

---

# Creating Data Frames

two dimensional arrays that can multiple classes of values.

  - syntax: *dataFrameSyntax <- data.frame(column1, column2, column3)*
    
```{r data frame, eval=FALSE}

#LINE 437

 title <- c("Star Wars", "The Empire Strikes Back", "Return of the Jedi")
  year <- c(1972, 1980, 1983)
  length.min <- c(121, 124, 133)
  box.office.mil <- c(787, 534, 572)
  
  starWars.data <- data.frame(title, year, length.min, box.office.mil)
  starWars.data

```
 
---

# Cleaning Data

### Rename Variables 

Change the variable names in the data frame.

  - syntax: *data.frame(<new_name> = <old_name>, ...)*
  
```{r rename variables, eval=FALSE}

starWars.data <- data.frame(Title = title, Year = year, Length = length.min, Gross = box.office.mil)
  starWars.data

```

### Subsetting Variables

look at and/or work on individual variable in a data frame.

  - syntax: *<object_name>$<variable_name>*
  
```{r subset, eval=FALSE}

starWars.data$Year

```
  
  
### Changing Data Values

change how R interpret the values in a variable

```{r change value, eval=FALSE}

starWars.data$Year <-  as.character(starWars.data$Year)
str(starWars.data)

```

### Changing Value of an Observation
  
Indexing is a way that you can call a column, row, or specific position in a data frame. You can use indexing to pinpoint a specific value and change that value.

  - syntax: *data.frame[rownumber, columnnumber] = newvalue*

```{r index, eval=FALSE}

starWars.data [1, 2]

```

```{r index example, eval=FALSE}

  starWars.data [1, 2] = "1977"
    starWars.data 

```

### Create New Variables 

You can use observations in your data frame to create new variables or overwrite a current variable
  
  - syntax: *<new_variable> = <name of object>$<name of variable><change>*
    
```{r new variable, eval=FALSE}

starWars.data$Gross.billions = starWars.data$Gross / 1000
  starWars.data

```
  
### Append new variables

Create new variables with observations and append them to your data frame

```{r append, eval=FALSE}

    han <- c(18.7, 21.4, 16.3)
  
  starWars.data$Han <- han
  
  starWars.data

```

### Merge new observations

Create new observations and merge it into a data frame
  
```{r merge, eval=FALSE}

solo <- data.frame(Title = "Solo", Year = 2018, Length = 135, Gross = 393.2, Gross.billions = 0.392,  Han = 0)

rbind(starWars.data, solo)

```

---

# Exporting data to a saved file

```{r export, eval=FALSE}

write.csv(starWars.data, "starwars.csv")

```

---

# Reading Data in R

### Base R option 

  - **read.table(...) / read.csv(...)**
 
```{r read, eval=FALSE}

help("read.table")

```

read.table will provide you with a generic way to load data where you are setting most of the parameters. However, the most common file for data frames are .csv files. R provides the *read.csv() function* which allows you to read .csv files with parameters set to default.

```{r read csv, eval=FALSE}

avengers <- read.csv("avengers.csv")
str(avengers)

```

### Tidyverse option 

  - **read_table(...) / read_csv(...)**
  
Using the tidyverse package you can read the data into a lighter frame.

```{r read tidyverse, eval=FALSE, message=FALSE}

fight_songs <- read_csv("fight-songs.csv")
fight_songs

```

---

# Exploring data

### Function: View()

view the full data frame
  
```{r View, eval =FALSE}

View(fight_songs)


# There is also a tidyverse option
  # view(avengers)

```

### Function: nrow()

view the number of rows/observations

```{r nrow, eval =FALSE}

    nrow(fight_songs)

```

### Function: ncol() 

view the number of columns/variables
    
```{r ncol, eval=FALSE}


  ncol(fight_songs)

```
  
### Function: colnames()

view the names of columns/variables
    
```{r colnames, eval=FALSE}

  colnames(fight_songs)

```

### Function: str()

view the structure of data
    
```{r structure, eval=FALSE}

  str(fight_songs)

```

### Function: summary()

basic descriptive statistics
    
```{r summary, eval=FALSE}


  summary(fight_songs)

```

### Function: mean(), median(), mode(), count(), sum() 

individual descriptive statistics
  
```{r descriptive statistics, eval=FALSE}

sum(fight_songs$number_fights)

```

### Function: head() or tail()

view top and bottom of data
    
```{r head and tail, eval=FALSE}


  head(fight_songs)

```

---

# Explore and Manipulate Data more with Tidyverse

```{r tidyverse, eval=FALSE}

library(tidyverse)
view(fight_songs)

```
  
### Pipe Operator - %>% 
  
The pipe operator allows you to separate and unite individual operations on an object
  
  - CTRL + SHIFT + M (windows)
  - CMD + SHIFT + M (mac)
        
*data %>% operation A %>% operation B*

### Function: arrange()

sort columns by values
  
```{r sort, eval=FALSE}
 
arrange(fight_songs, desc(conference))  ## Most / z-a
arrange(fight_songs, conference) ## Least / a-z

```

### Function: filter()

limit observations in a variable based on criteria
  
```{r filter, eval=FALSE}

fight_songs_big12 <- fight_songs %>% filter(conference == "Big 12")
mean(fight_songs$bpm)
mean(fight_songs_big12$bpm)

```

### Function: select()

limit the number of variables in the data frame

```{r select, eval =FALSE}

songs_by_school <- fight_songs %>% select(school, song_name)
songs_by_school

```

### Function: groupby() and summarise

create dummy variables for categorical variables and get summary statistics
  
  
```{r group by, eval=FALSE}

conference_num_fights <- fight_songs %>% 
  group_by(conference) %>% 
  summarise (avg_fight = mean(number_fights))

conference_num_fights

```

